---
title: "Homework 1 cahpter2"
author: "Guibril Ramde"
format:
  html: default
  pdf: default
---

## 2.10 Exercises

**2.1**

```{r}
library(fpp3)
library(tsibble)
help(aus_production)
# A tsibble: 218 x 7 [1Q]
  # Quarter  Beer Tobacco Bricks Cement Electricity   Gas
 #    <qtr> <dbl>   <dbl>  <dbl>  <dbl>       <dbl> <dbl>
 #1 1956 Q1   284    5225    189    465        3923     5
# 2 1956 Q2   213    5178    204    532        4436     6
# 3 1956 Q3   227    5297    208    561        4806     7
# 4 1956 Q4   308    5681    197    570        4418     6


help(vic_elec)
# A tsibble: 52,608 x 5 [30m] <Australia/Melbourne>
  # Time                Demand Temperature Date       Holiday
  # <dttm>               <dbl>       <dbl> <date>     <lgl>  
# 1 2012-01-01 00:00:00  4383.        21.4 2012-01-01 TRUE   
# 2 2012-01-01 00:30:00  4263.        21.0 2012-01-01 TRUE   
# 3 2012-01-01 01:00:00  4049.        20.7 2012-01-01 TRUE   

help(pelt)
# A tsibble: 91 x 3 [1Y]
   # Year  Hare  Lynx
 #  <dbl> <dbl> <dbl>
# 1  1845 19580 30090
# 2  1846 19600 45150
# 3  1847 19610 49150
# ℹ 81 more rows

help(gafa_stock)
# A tsibble: 5,032 x 8 [!]
# Key:       Symbol [4]
   #Symbol Date        Open  High   Low Close Adj_Close    Volume
   #<chr>  <date>     <dbl> <dbl> <dbl> <dbl>     <dbl>     <dbl>
# 1 AAPL   2014-01-02  79.4  79.6  78.9  79.0      67.0  58671200
# 2 AAPL   2014-01-03  79.0  79.1  77.2  77.3      65.5  98116900
# ℹ 5,022 more rows
```

> Bricks in aus_production is a quarterly data

> Demand from vic_elec is half hourly data
>
> Lynx from pelt is a yearly data
>
> Close from gafa_stock is daily data

```{r}
aus_production |>
  autoplot(Bricks, na.rm =TRUE)
  

```

```{r}
vic_elec |>
  autoplot(Demand)





```

```{r}
gafa_stock |>
  autoplot(Close)
```

```{r}
pelt |>
  autoplot(Lynx)

```

> Modify the last plot axis label and title

```{r}
vic_elec |>
  autoplot(Demand) +
  labs(
    title = "Electricity Demand in Victoria",
    x = "Date and Time",
    y = "Electricity Demand"
  )

```

**2.2**

```{r}
gafa_stock |>
  group_by(Symbol) |>
  filter(Close == max(Close)) |>
  select(Symbol, Date, Close)

```

**2.3**

```{r}
load_data <- read.csv("tute1.csv")
load_data
```

a\. read the data into R

```{r}
tute1 <- readr::read_csv("tute1.csv")
view(tute1)
```

b\. convert the data to time series

```{r}
mytimeseries <- tute1 |>
  mutate(Quarter = yearquarter(Quarter)) |>
  as_tsibble(index = Quarter)
```

c\. construct time series plots of each of the three series

```{r}
mytimeseries |>
  pivot_longer(-Quarter) |>
  ggplot(aes(x = Quarter, y = value, colour = name)) +
  geom_line() +
  facet_grid(name ~ ., scales = "free_y")
```

not including facet_grid the 03 name get store in the same value making it one plot, which might be confusing during interpretation.

**2.4**

```{r}
library(USgas)
names(us_total)
```

```{r}
#install.packages("USgas")

us_total <- us_total |>
  as_tsibble(key = state, index =  year)
us_total
is_tsibble(us_total)
```

c\. Plot the annual natural gas

```{r}
new_england <- c(
   "Maine",
  "Vermont",
  "New Hampshire",
  "Massachusetts",
  "Connecticut",
  "Rhode Island"
  
)
us_total |>
  filter(state %in% new_england) |>
  autoplot(y)
```

**2.5 a. read tourism.xlsx**

```{r}
load_data_t <- readxl::read_excel("tourism.xlsx")
load_data_t
```

b.create tsibble

```{r}
tourism_data <- load_data_t |>
  mutate(Quarter = yearquarter(Quarter)) |>
  as_tsibble(key = c(Region,State, Purpose), index = Quarter)
tourism_data

```

```{r}
#checking if it is identical 
tourism
key_vars(tourism)
index_var(tourism)
```

c**. Max overnight trip**

```{r}
max_trip <- tourism_data |>
  group_by(Region, Purpose) |>
  summarise(avg_trips = mean(Trips), .groups = "drop") |>
  filter(avg_trips == max(avg_trips))
  

max_trip
```

d.Create a new tsibble which combine region and purpose

```{r}
tourism_new <- tourism_data |>
  group_by(State) |>
  index_by(Quarter) |>
  summarise(Trips = sum(Trips))

tourism_new
```

**2.8 Use the following graphics functions**

```{r}
total_private <- us_employment |>
  filter(Title == "Total Private")

total_private |> autoplot(Employed)
total_private |> gg_season(Employed)
total_private |> gg_subseries(Employed)
total_private |> gg_lag(Employed)
total_private |> ACF(Employed) |> autoplot()
```

-   Observation: **Trend:** There is a strong increasing trend in private employment over time. More recent years have much higher employment levels than earlier years. Cause could be internet, technologies, immigration. **Cyclicity:** There are periods where employment decreases for some time before recovering and continuing its long-term growth, suggesting cyclical movements, the autoplot is better to visualize that cyclicity pattern.

-   Private employment has a strong upward trend over time, although there are some periods of decline and recovery. This due to school ending, summer jobs, entry - level jobs and also holidays part time jobs.

-   Employment generally increases from January to June, remains relatively stable during the summer, decreases slightly around September, and then slightly increases or remains stable toward the end of the year. This pattern repeats across many years.

-   Some years show larger decreases than the usual seasonal pattern. The time-series plot shows particularly noticeable downturns around 2008–2009 and 2020.

```{r}
aus_production |>
  ACF(Bricks, na.rm = TRUE) |> autoplot()

aus_production |>
  autoplot(Bricks, na.rm = TRUE)

aus_production |>
  gg_subseries(Bricks, na.rm = TRUE) 

aus_production |>
  gg_season(Bricks, na.rm = TRUE)

aus_production |>
  gg_lag(Bricks, geom = "point")

```

**Observation:**

-   Observation: **Trend:** There is a strong increasing trend on aus_productions 'Bricks' over time. More recent years have much higher Bricks production than earlier years. Cause could be population growth, immigration, housing, tourism. **Cyclicity:** There are periods where Bricks production decreases for some time before recovering and continuing its long-term growth, suggesting cyclical movements, the autoplot is better to visualize that cyclicity pattern.

-   Bricks production has a strong upward trend over time, although there are some periods of decline and recovery like around years 1980-1984 which we can observe a peak. the same pattern repeat over the time.

-   Bricks production generally increases from Q1 to Q2, remains relatively stable from Q2 to Q3 then decreases slightly from Q3 to Q4.

-   1975 to 1985 are the years we see increase on Bricks production in Australia, then down during early 2000's to 2005.

```{r}
pelt |>
  autoplot(Hare, na.rm = TRUE)

pelt |>
  gg_lag(Hare, geom = "point")

 #pelt |>
  #gg_season(Hare, na.rm = TRUE)
 
# pelt |>
 # gg_subseries(Hare, na.rm = TRUE)
 
 pelt |>
  ACF(Hare, na.rm = TRUE) |> autoplot()
 
  
```

-   Observation: **Trend:** There is a strong increasing and decrese trend in Hares population over time. **Cyclicity:** The pattern tend to repeat throughout the years, increase - decrease and not stable.

-   The series show how the Hares population are affected from 1800 to 1920 with a pick in 1865 right followed with a strong decrease and pattern is repeated.

-   For this Hare data we could not use a seasonal plot because it is reported yearly.

-   Some cycles have much larger peaks than others, particularly around the 1860s and 1880s. noticeable downturns around 1863, 1868– 1900 and 1920 also a upturn increase in years 1865 and 1885 .

```{r}
h02 <- PBS |>
  filter(ATC2 == "H02") |>
  group_by(ATC2) |>
  index_by(Month) |>
  summarise(Cost = sum(Cost))

h02 |>
  gg_lag(Cost, geom = "point")
h02 |>
  gg_season(Cost)
h02 |>
  gg_subseries(Cost)
h02 |>
  autoplot(Cost)
 h02 |>
   ACF(Cost) |> autoplot()
```

-   Observation: **Trend:** There is a strong decrease from January to February, then slightly upward increase from February through December trend in Cost from PBS over time. The same pattern repeat throughout the years 1995 to 2005. **Seasonality:** There are periods where Cost decreases for some time before recovering and continuing its long-term growth, suggesting cyclical movements, the seasonal plot is better to visualize that cyclicity pattern.

-   Cost has a strong upward trend over time, although there are some periods of decline and recovery. This due to new year and holidays.

-   Cost generally increases relatively from February through December. This pattern repeats across many years.

-   unusual years I can identify is year 2005 where we can observe a decrease from February through March

```{r}
us_gasoline |>
  gg_season(Barrels)

us_gasoline |>
  gg_subseries(Barrels)

us_gasoline |>
  gg_lag(Barrels, geom = "point")

us_gasoline |>
  autoplot(Barrels)

us_gasoline |>
  ACF(Barrels) |> autoplot()
 
```

-   Observation: **Trend:** There is a strong increasing trend in USA Barrels over time. More recent years have much higher number of Barrels than earlier years. **Cyclicity:** There are periods where the number decreases for some time before recovering vice versa, suggesting cyclical movements, the autoplot is better to visualize that cyclicity pattern.

-   Barrels number has a strong upward trend over time, although there are some periods of decline and recovery. This due to the USA importing Barrels and competition.

-   The number of Barrels has increase over time, observing the seasonal pattern we see an up and down.

-   The trends are almost the same through years 1995 to 2015 almost the same pattern occur every year.
