---
title: "chapter2_homeworks"
format: html
editor: visual
---

## 2.10 Exercises

2.1

```{r}
library(fpp3)
library(tsibble)
help(aus_production)
# A tsibble: 218 x 7 [1Q]
  # Quarter  Beer Tobacco Bricks Cement Electricity   Gas
 #    <qtr> <dbl>   <dbl>  <dbl>  <dbl>       <dbl> <dbl>
 #1 1956 Q1   284    5225    189    465        3923     5
# 2 1956 Q2   213    5178    204    532        4436     6
# 3 1956 Q3   227    5297    208    561        4806     7
# 4 1956 Q4   308    5681    197    570        4418     6


help(vic_elec)
# A tsibble: 52,608 x 5 [30m] <Australia/Melbourne>
  # Time                Demand Temperature Date       Holiday
  # <dttm>               <dbl>       <dbl> <date>     <lgl>  
# 1 2012-01-01 00:00:00  4383.        21.4 2012-01-01 TRUE   
# 2 2012-01-01 00:30:00  4263.        21.0 2012-01-01 TRUE   
# 3 2012-01-01 01:00:00  4049.        20.7 2012-01-01 TRUE   

help(pelt)
# A tsibble: 91 x 3 [1Y]
   # Year  Hare  Lynx
 #  <dbl> <dbl> <dbl>
# 1  1845 19580 30090
# 2  1846 19600 45150
# 3  1847 19610 49150
# ℹ 81 more rows

help(gafa_stock)
# A tsibble: 5,032 x 8 [!]
# Key:       Symbol [4]
   #Symbol Date        Open  High   Low Close Adj_Close    Volume
   #<chr>  <date>     <dbl> <dbl> <dbl> <dbl>     <dbl>     <dbl>
# 1 AAPL   2014-01-02  79.4  79.6  78.9  79.0      67.0  58671200
# 2 AAPL   2014-01-03  79.0  79.1  77.2  77.3      65.5  98116900
# ℹ 5,022 more rows
```

> Bricks in aus_production is a quarterly data

> Demand from vic_elec is half hourly data
>
> Lynx from pelt is a yearly data
>
> Close from gafa_stock is daily data

```{r}
aus_production |>
  autoplot(Bricks)
  

```

```{r}
vic_elec |>
  autoplot(Demand)





```

```{r}
gafa_stock |>
  autoplot(Close)
```

```{r}
pelt |>
  autoplot(Lynx)

```

> Modify the last plot axis label and title

```{r}
vic_elec |>
  autoplot(Demand) +
  labs(
    title = "Electricity Demand in Victoria",
    x = "Date and Time",
    y = "Electricity Demand"
  )

```

2.2

```{r}
gafa_stock |>
  group_by(Symbol) |>
  filter(Close == max(Close)) |>
  select(Symbol, Date, Close)

```

2.3

```{r}
load_data <- read.csv("tute1.csv")
load_data
```

a\. read the data into R

```{r}
tute1 <- readr::read_csv("tute1.csv")
view(tute1)
```

b\. convert the data to time series

```{r}
mytimeseries <- tute1 |>
  mutate(Quarter = yearquarter(Quarter)) |>
  as_tsibble(index = Quarter)
```

c\. construct time series plots of each of the three series

```{r}
mytimeseries |>
  pivot_longer(-Quarter) |>
  ggplot(aes(x = Quarter, y = value, colour = name)) +
  geom_line() +
  facet_grid(name ~ ., scales = "free_y")
```

not including facet_grid the 03 name get store in the same value making it one plot, which might be confusing during interpretation.

2.4

```{r}
names(us_total)
```

```{r}
#install.packages("USgas")
library(USgas)
us_total <- us_total |>
  as_tsibble(key = state, index =  year)
us_total
is_tsibble(us_total)
```

c\. Plot the annual natural gas

```{r}
new_england <- c(
   "Maine",
  "Vermont",
  "New Hampshire",
  "Massachusetts",
  "Connecticut",
  "Rhode Island"
  
)
us_total |>
  filter(state %in% new_england) |>
  autoplot(y)
```
