---
title: "forecaster's toolbox"
author: "Guibril Ramde"
format:
  html: default
  pdf: default
editor: visual
---

## Exercise 5

##  5.1 Produce forecasts for the following series using whichever of `NAIVE(y)`, `SNAIVE(y)` or `RW(y ~ drift())` is more appropriate in each case:

```{r}
library(fpp3)
library(tsibble)
library(tidyverse)
library(dplyr)
library(ggplot2)
#install.packages("fpp3")
library(ggtime)
```

## **Australian Population (global economy)**

```{r}
#Data preparation
aus_population <- global_economy |>
  filter(Country == "Australia")
# Plot visualization
aus_population |>
  autoplot(Population) +
  labs(
    y = "Population",
    title = "Australian Population"
  )

#Model selection
fit_population <- aus_population |>
  model(RW(Population ~ drift()))
#generate and forecast in 3 years
population_fc <- fit_population |>
  forecast(h = "3 years")
#plot the forecast
population_fc |>
  autoplot(aus_population)
```

## **Interpretation:**

Australian population shows a clear upward trend and no seasonal pattern. Therefore, I selected the random walk with drift model.

**#Bricks (aus_production)**

```{r}
# Data preparation (tidy) and Plot (visualization)
aus_prod <- aus_production |>
  select(Bricks)
bricks_fit <- aus_prod |>
  model(SNAIVE(Bricks))

bricks_fc <- bricks_fit |>
  forecast(h = 12)

bricks_fc |>
  autoplot(aus_prod, level = NULL) +
  labs(
    y = "Bricks",
    title = "Forecast for Quarterly Bricks Production"
  )

```

## **Interpretation:**

Since bricks is a quarterly data and a strong seasonality a SNAIVE is fit to use.

**##NSW Lambs (aus_livestock)**

```{r}
aus_livestock |>
  distinct(State)

nsw_lambs <- aus_livestock |>
  filter(State == "New South Wales",  Animal == "Lambs")
nsw_lambs
nsw_lambs |>
  autoplot(Count)
#Model selection
fit_nsw <- nsw_lambs |> model(SNAIVE(Count))
#Generate forecast
nsw_fit <- fit_nsw |> forecast(h = 8)
#Plot forecast
nsw_fit |>
  autoplot(nsw_lambs, level = NULL) +
  #autolayer(filter_index(nsw_lambs, "2020" ~ .), colour = "red") +
  labs(title = "New South Wales Lambs")

    

```

## **Interpretation:**

The NSW Lambs series shows a seasonal pattern, so the seasonal naïve method is appropriate.

**\## Household Wealth**

```{r}
hh_budget |>
  autoplot(Wealth)

fit_house <- hh_budget |>
  model(RW(Wealth ~ drift()))

house_fc <- fit_house |>
  forecast(h = 10)

house_fc |>
  autoplot(hh_budget, level = NULL) +
  labs(title = "Household Wealth")
```

## **Interpretation:**

RW model is good fit for the forecast of this datasets because the models makes sense for a series dominated by a trend without a strong seasonal pattern.

## **Retail**

```{r}
aus_retail |>
  distinct(State)
aus_retail |>
  distinct(Industry)

food_turnover <- aus_retail |>
  filter(Industry == "Cafes, restaurants and takeaway food services") |>
  index_by(Month) |>
  summarise(Turnover = sum(Turnover))

food_turnover |>
  autoplot(Turnover) +
  labs(
    title = "Australian Cafes, Restaurants and Takeaway Food Turnover",
    y = "Turnover"
  )
# Model selection
fit_food <- food_turnover |> model(SNAIVE(Turnover))

# Generate forecast
fit_fd <- fit_food |> forecast(h = 12)
fit_fd |>
  autoplot(food_turnover, level = NULL ) +
  #autolayer(filter_index(aus_retail, '2020' ~ .), colour = 'red') +
  labs(
    title = "Australian Takeaway Food Turnover Forecast",
    y = "Turnover"
  )
  
```

## **Interpretation:**

The Australian takeaway food turnover series shows a recurring seasonal pattern. Therefore, I selected the seasonal naïve method, which uses the corresponding observation from the previous year to forecast each month.

## **5.2 Use the Facebook stock price (data set `gafa_stock`) to do the following:**

a. Produce a time plot of the series

```{r}
fb_data <- gafa_stock |>
  filter(Symbol == "FB") |>
  mutate(day = row_number()) |>
  update_tsibble(index = day, regular = TRUE)
fb_data |>
  autoplot(Close) +
  labs(
    title = "Facebook Stock Price",
    y = "Closing Price"
  )
#b Produce Forecast using drift
fit_fb <- fb_data |> model(RW(Close ~ drift()))
fit_fb
#forecast
fb_fit <- fit_fb |> forecast(h = 200)
#Plot the forecast
fb_fit |>
  autoplot(fb_data) +
  labs(
    title = "Facebook Stock Price Forecast",
    y = "Closing Price",
    x = "Trading Day"
  )
```

## **C. Show that the forecasts are identical to extending the line drawn between the first and last observations.**

```{r}
first_price <- first(fb_data$Close)
first_price
last_price <- last(fb_data$Close)
last_price
n <- nrow(fb_data)
n

#Calculating the slop
slope <- (last_price - first_price)/(n - 1)
slope

##Calculating future values
future_days <- 1:200

extend_line_price <- last_price + future_days * slope
extend_line_price


```

## **Part D Try using some of the other benchmark functions to forecast the same data set. Which do you think is best? Why?**

```{r}
fb_fit <- fb_data |>
  model(
    Naive = NAIVE(Close),
    Drift = RW(Close ~ drift()),
    Mean = MEAN(Close)
  )
#Forecast
forecast_fb <- fb_fit |> forecast(h = 200)
forecast_fb |>
  autoplot(fb_data)
```

## **Interpretation:**

Looking at the forecasts, I think the drift method is more appropriate. The Facebook stock price shows an overall increasing trend over the observed period. The mean method does not account for this trend, while the naïve method assumes that future prices will remain at the last observed value. The drift method accounts for the overall upward movement by extending the average change between the first and last observations.

## **5.3. Apply a seasonal naïve method to the quarterly Australian beer production data from 1992. Check if the residuals look like white noise, and plot the forecasts. The following code will help.**

```{r}
# Extract data of interest
recent_production <- aus_production |>
  filter(year(Quarter) >= 1992)
# Define and estimate a model
fit <- recent_production |> model(SNAIVE(Beer))
# Look at the residuals
fit |> gg_tsresiduals()
# Look a some forecasts
fit |> forecast() |> autoplot(recent_production)
```

## **Interpretation:**

The residuals do not appear to be completely white noise. The ACF plot shows a significant negative spike at lag 4 and a positive spike around lag 3 that exceeds the significance bounds. Since the data are quarterly, lag 4 represents one year, suggesting that some seasonal information remains in the residuals. Therefore, the seasonal naïve model captures the general seasonal pattern, but it does not capture all of the information in the series.

## **5.4 Repeat the previous exercise using the Australian Exports series from `global_economy` and the Bricks series from `aus_production`. Use whichever of `NAIVE()` or `SNAIVE()` is more appropriate in each case.**

```{r}
# Extract data of interest
australian_export <- global_economy |>
  filter(Country == "Australia")
# Define and estimate a model
fit <- australian_export |> model(NAIVE(Exports))
# Look at the residuals
fit |> gg_tsresiduals()
# Look a some forecasts
fit |> forecast() |> autoplot(australian_export)
```

```{r}
# Extract data of interest
australian_bricks <- aus_production |>
  filter(year(Quarter) >= 1970) |>
  filter(!is.na(Bricks))
# Define and estimate a model
fit_brick <- australian_bricks |> model(SNAIVE(Bricks))
# Look at the residuals
fit_brick |> gg_tsresiduals()
# Look a some forecasts
fit_brick |> forecast() |> autoplot(australian_bricks)
```

## **Interpretation:**

For the Australian Exports series, I used the naïve method because the data are annual and do not have a seasonal pattern. The naïve method forecasts future exports using the most recent observed value. For the Bricks series, I used the seasonal naïve method because the quarterly data show a recurring seasonal pattern. The seasonal naïve method forecasts each quarter using the value from the corresponding quarter of the previous year.

The residuals for both datasets do not appear to be completely white noise. For Australian Exports, the ACF plot shows a significant negative spike at lag 1, indicating that some autocorrelation remains in the residuals. For Bricks production, the ACF plot shows several significant positive and negative spikes that exceed the significance bounds. This indicates that there is still autocorrelation in the residuals. Therefore, although the seasonal naïve method captures the general seasonal pattern in Bricks production, it does not capture all of the information in the series.

**5.7 For your retail time series (from Exercise 7 in Section [2.10](https://otexts.com/fpp3/graphics-exercises.html#graphics-exercises)):**

```{r}
set.seed(12345678)
myseries <- aus_retail |>
  filter(`Series ID` == sample(aus_retail$`Series ID`,1))

myseries_train <- myseries |>
  filter(year(Month) < 2011)

autoplot(myseries, Turnover) +
  autolayer(myseries_train, Turnover, colour = "red")

fit <- myseries_train |>
  model(SNAIVE(Turnover))

fit |> gg_tsresiduals()

fc <- fit |>
  forecast(new_data = anti_join(myseries, myseries_train))
fc |> autoplot(myseries)

fit |> accuracy()
fc |> accuracy(myseries)

```

## **g. How sensitive are the accuracy measures to the amount of training data used?**

```{r}
train1 <- myseries |>
  filter(year(Month) >= 1990, year(Month) < 2011)


train2 <- myseries |>
  filter(year(Month) >= 2000, year(Month) < 2011)

train3 <- myseries |>
  filter(year(Month) >= 2005, year(Month) < 2011)



fit1 <- train1 |>
  model(SNAIVE(Turnover))


fit2 <- train2 |>
  model(SNAIVE(Turnover))


fit3 <- train3 |>
  model(SNAIVE(Turnover))


test <- myseries |>
  filter(year(Month) >= 2011)

fc1 <- fit1 |> forecast(new_data = test)
fc2 <- fit2 |> forecast(new_data = test)
fc3 <- fit3 |> forecast(new_data = test)

fc1 |> accuracy(myseries)
fc2 |> accuracy(myseries)
fc3 |> accuracy(myseries)
#range(myseries$Month)
```

## **Interpretation:**

I tested the seasonal naïve model using different amounts of training data by starting the training period in 1990, 2000, and 2005, while keeping the end of the training period at 2010. The test accuracy measures were identical for all three training sets, with an RMSE of 1.552, MAE of 1.241, and MAPE of 9.064%. Therefore, the forecast accuracy of the seasonal naïve method is not sensitive to the amount of older training data used in this case. This occurs because the seasonal naïve method bases its forecasts on the most recent seasonal observations, which are the same for all three training sets.
